---
title: "Assignment_5"
author: "Samuel Estrada Jasso"
date: "2025-05-30"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: true
    toc_depth: 3
---


```{r, echo=FALSE, include=FALSE}
library(knitr)
library(tidyverse)
library(discrtr)
library(mlogit)
library(readr)
library(plotly)
library(gplots)
library(ggmosaic) 
library(treemapify)
library(ggridges)
library(ggalluvial)
library(dplyr)
library(ggplot2)
library(stargazer) 
library(evd)
library(htmlwidgets)
library(kableExtra)
library(tidyr)
library(webshot2)
library(dfidx)
```

# **Question 1.**
What is the difference between a marginal effect and an elasticity?

<p style="color:green;">The marginal effect measures how much the dependent variable changes when an independent variable $x_k$ increases by one unit. In contrast, elasticity indicates how the dependent variable proportionally varies in response to percentage changes in $x_k$.</p>

# **Question 2.**
Why is it not appropriate to calculate the elasticity of a dummy variable?

<p style="color:green;">Marginal effects and elasticities are not suitable for qualitative (dummy) variables since it does not make sense to discuss unitary or percentage increases in categories like "mountain region."</p>

# **Question 3.** 
Use Model 3 from this chapter and calculate the marginal effects and elasticities of the operating cost at the mean of all variables.


```{r}
data("Heating")

H <- Heating %>% mlogit.data(shape = "wide", choice="depvar", varying=c(3:12))

model3 <- mlogit(depvar ~ ic + oc, data = Heating, shape = "wide", choice = "depvar", reflevel = "ec", varying = 3:12
)

medias <- Heating[1, ]

mean_cost <- Heating %>%
  select(starts_with("ic"), starts_with("oc")) %>%
  summarise(across(.cols = everything(), mean))

medias[3:12] <- mean_cost

efectos_marginales <- effects(model3, covariate = "oc", type = "ra", data = mlogit.data(medias, shape = "wide", choice = "depvar", varying = 3:12))
print("Efectos marginales:")
print(efectos_marginales)

elasticidades <- effects(model3, covariate = "oc", type = "rr", data = mlogit.data(medias, shape = "wide", choice = "depvar", varying = 3:12))
print("Elasticidades:")
print(elasticidades)
```

# **Question 4.** 
Use Model 3 from this chapter to calculate the reimbursement needed to achieve a 10% penetration rate of heat pumps.

```{r}
rebates <- seq(0, 2000, by = 50)

hp_shares <- numeric(length(rebates))

Heating_original <- Heating

for (i in seq_along(rebates)) {
  Heating_temp <- Heating_original
  Heating_temp$ic.hp <- Heating_temp$ic.hp - rebates[i]
  
  Heating_long <- mlogit.data(Heating_temp, shape = "wide", choice = "depvar",
                              varying = 3:12, id.var = "idcase")
  
  prob_temp <- predict(model3, newdata = Heating_long, type = "probabilities")
  
  hp_shares[i] <- mean(prob_temp[, "hp"])
}

target_share <- 0.10
required_rebate <- rebates[min(which(hp_shares >= target_share))]
required_rebate
```

<p style="color:green;">A cost reduction of 450 monetary units is required to reach the desired quota.</p>

## **Question 4.1** 
Now, estimate a new extended model based on Model 3, including the age of the household head as an explanatory variable, and use the room electrical system ("er") as the reference category.


```{r}
model4 <- mlogit(depvar ~ ic + oc | agehed, Heating, shape = "wide", choice = "depvar", reflevel = "er", varying = c(3:12))

summary(model4)
```

# **Question 5** 
Use the likelihood ratio test to compare your new model with Model 3. Discuss the results.

```{r}
lrtest(model3, model4)
```

<p style="color:green;">Including the variable *agehed* increases the likelihood ratio, but the difference is not statistically significant. This indicates that the age of the household head does not have a relevant influence on the choice of heat pumps in this case.</p> 

# **Question 6** 
Is the ratio between the installation (or operating) cost coefficient and the household head's age coefficient significant? Explain.

<p style="color:green;">It is not valid to interpret both ratios since they correspond to different types of variables—one specific to the alternative and the other to the individual.</p>

