---
title: "Assignment_6"
author: "Samuel Estrada Jasso"
date: "2025-05-30"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: true
    toc_depth: 3
---


```{r, echo=FALSE, include=FALSE}
library(knitr)
library(tidyverse)
library(discrtr)
library(mlogit)
library(readr)
library(plotly)
library(gplots)
library(ggmosaic) 
library(treemapify)
library(ggridges)
library(ggalluvial)
library(dplyr)
library(ggplot2)
library(stargazer) 
library(evd)
library(htmlwidgets)
library(kableExtra)
library(tidyr)
library(webshot2)
library(dfidx)
```

# **Exercise 2**
Use model nl2 in this chapter and calculate the direct-point elasticity at the mean values of the variables, for an increase in the installation costs of Gas Central systems.

```{r}
data("Heating", package = "mlogit")

H <- mlogit.data(Heating, shape = "wide", choice = "depvar", varying = c(3:12))

nl2 <- mlogit(depvar ~ ic + oc, H, nests = list(room = c( 'er', 'gr'), central = c('ec', 'gc', 'hp')), un.nest.el = TRUE, steptol = 1e-12)

gc_data <- subset(H, alt == "gc")

mean_ic_gc <- mean(gc_data$ic)

probs <- fitted(nl2, type = "probabilities")

mean_prob_gc <- mean(probs[H$alt == "gc"])

beta_ic <- coef(nl2)["ic"]

elasticity_gc_ic <- beta_ic * mean_ic_gc * (1 - mean_prob_gc)
elasticity_gc_ic
```

<p style="color:green;">A 1% increase in the installation cost of the Central Gas system decreases the probability of being chosen by 0.052%.</p>

# **Exercise 3**
Use model nl2 in this chapter and calculate the cross-point elasticity at the mean values of the variables, for a 1% increase in the operation costs of Gas Central systems.

```{r}
gc_data <- subset(H, alt == "gc")

mean_oc_gc <- mean(gc_data$oc)

probs <- fitted(nl2, type = "probabilities")

mean_probs <- tapply(probs, H$alt, mean)

beta_oc <- coef(nl2)["oc"]

P_ec <- mean_probs["ec"]
P_gc <- mean_probs["gc"]

elasticity_cross_ec_gc <- -beta_oc * mean_oc_gc * (P_gc / P_ec)
elasticity_cross_ec_gc
```

<p style="color:green;">A 1% increase in the operating cost of the Central Gas system increases the probability of consumers choosing other options by 0.039%.</p>

# **Exercise 4**
Re-estimate the nested logit model in this chapter, but change the nests to types of energy as follows:

Gas: gas central, gas room.

Electricity: electric central, electric room, heat pump.

Use a single coefficient for the inclusive variables (i.e.,set un.nest.el = TRUE). Are the results reasonable? Discuss.

```{r}
nests_energy <- list(
  gas = c("gc", "gr"),        
  electricity = c("ec", "er", "hp")
)

nl_energy <- mlogit(depvar ~ ic + oc, data = H,
                    nests = nests_energy,
                    un.nest.el = TRUE)

summary(nl_energy)
```

<p style="color:green;">The installation and operating cost coefficients are negative and significant, suggesting that higher costs reduce perceived utility.</p>

