---
title: "Assignment_4"
author: "Samuel Estrada Jasso"
date: "2025-05-28"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: true
    toc_depth: 3
---

```{r, echo=FALSE, include=FALSE}
library(knitr)
library(tidyverse)
library(discrtr)
library(mlogit)
library(readr)
library(plotly)
library(gplots)
library(ggmosaic) 
library(treemapify)
library(ggridges)
library(ggalluvial)
library(dplyr)
library(ggplot2)
library(stargazer) 
library(evd)
library(htmlwidgets)
library(kableExtra)
library(tidyr)
library(webshot2)
library(dfidx)
```

# **Exercise 1**
In the example in this chapter we estimated the probabilities of choosing different modes by sidewalk density setting travel time to the in-sample median. Use model2 to calculate the probability of choosing different modes by in-sample median sidewalk density but now for travel times of 20, 30, and 40min. Discuss the results. 

```{r}
data("mc_commute_wide", package = "discrtr")

mc_commute_long <- mc_commute_wide |> 
mlogit.data(shape="wide",
            choice = "choice",
            varying = 3:22)

mc_commute_predict <- mc_commute_long[1:52, ]
mc_commute_predict$sidewalk_density <- rep(seq(from=0, to=60, by=5), each=4)
```

```{r}
f2 <- mFormula(choice ~ time | sidewalk_density)

model2 <- mlogit(f2, mc_commute_long)

times <- c(20, 30, 40)
results <- list()

for (t in times) {
    mc_commute_predict$time <- t
    probs <- predict(model2, newdata=mc_commute_predict)
    results[[as.character(t)]] <- probs
}

for (t in times) {
    cat("\nProbabilities for travel time =", t, "minutes:\n")
    print(results[[as.character(t)]])
}
```

<p style="color:green;">At this point, $f2$ assumes that travel time affects the total number of options in the same proportion, and as travel time increases, the utility of each alternative decreases by the same amount.
</p>

# **Exercise 2**
Estimate a model using formula f3 (call it model3). Discuss the output of this model.

```{r}
f3 <- mFormula(choice ~ 0 | sidewalk_density | time)

model3 <- mlogit(f3, mc_commute_long)

summary(model3)
```

<p style="color:green;">On the contrary, model3 allows travel time to affect each mode of transportation differently.</p>


# **Exercise 3**
Use model3 to calculate the probability of choosing different modes by in sample median sidewalk density but now for travel times of 20, 30, and 40min. Discuss the results.

```{r}
times <- c(20, 30, 40)
results <- list()

for (t in times) {
    mc_commute_predict$time <- t
    probs <- predict(model3, newdata=mc_commute_predict)
    results[[as.character(t)]] <- probs
}

for (t in times) {
    cat("\nProbabilities for travel time =", t, "minutes:\n")
    print(results[[as.character(t)]])
}
```
<p style="color:green;">The longer the travel time, the lower the probability of walking or choosing a car or bus. On the contrary, the probability of opting for a bicycle increases.</p>


# **Exercise 4**
In the general case, what is the value of the log-likelihood of the null (Equally Likely) model?

```{r}
f0 <- mFormula(choice ~ 1)
model0 <- mlogit(f0, mc_commute_long)
summary(model0)
```

# **Exercise 5**
Use the likelihood ratio test to compare model3 to model2? Discuss the results. What restrictions would you need to impose in model3 to obtain model2?

```{r}
lrtest(model2, model3)
```

<p style="color:green;">Model3 fits significantly better than Model2, suggesting that allowing the effect of time to vary by mode of transportation improves the model.</p>
<p style="color:green;">To obtain Model2, the time coefficient must be the same across all alternatives.</p>


