---
title: "Assignment 3"
author: "Samuel Estrada Jasso"
date: "2025-05-27"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: true
    toc_depth: 3
---

```{r, echo=FALSE, include=FALSE}
library(knitr)
library(tidyverse)
library(discrtr)
library(mlogit)
library(readr)
library(plotly)
library(gplots)
library(ggmosaic) 
library(treemapify)
library(ggridges)
library(ggalluvial)
library(dplyr)
library(ggplot2)
library(stargazer) 
library(evd)
library(htmlwidgets)
library(kableExtra)
library(tidyr)
library(webshot2)
library(dfidx)
```

# **Excercise 5**
Consider the following utility functions for two alternatives, namely i and j:

$U_i=V_i+ε_i$

$U_j=V_j+ε_j$

Assume that the difference between the error terms below follows the triangle distribution:

$ε_q=ε_i−ε_j$

Parting from the assumption above, derive a binary choice model for the probability of selecting alternative j.

```{r}
f_den <- function(x, B=1){
  if(x>=-B & x<=0){
    y <- 1/B^2 *x +1/B
  }else if(x>0 & x<=B){
    y <- -1/B^2 *x +1/B
}else{
    y <- 0
}
  return(y)
}

fv_den <- Vectorize(f_den, "x")

f_dis <- function(x, B=1){
  if(x<(-1)*B){
    y <- 0}else if(x>=-B & x<=0){
    y <- (x+B)*(fv_den(x)/2)
  }else if(x>0 & x<=B){
    y <- 1- (B-x)*fv_den(x)/2
} else{
    y <- 0
}
  return(y)
}

x <- seq(-1.5, 1.5, length.out=500)
y <- sapply(x, f_den)

ggplot()+
  geom_line(aes(x,y), color="blue", , linewidth=1.5)+
  geom_function(fun = fv_den, color="darkgreen", linewidth=0.5)+
  xlim(-1.5, 1.5)+
  theme_bw()

decision <- function(Vj, Vi, B=1){
  Vji <- Vj-Vi
  PVj <- f_dis(Vji)
  
  ggplot()+
  geom_function(fun = fv_den, color="blue", linewidth=0.5, xlim=c(-B-0.1, B+0.1))+
  stat_function(fun = fv_den, geom= "area", fill="darkgreen", alpha=0.5, xlim=c(-B, Vji))+ 
  geom_text(aes(B/2, 1/(2*B), label=paste0("Prob(Vj)=", round(PVj,2), sep="")))+
    labs(title="Probabilidad de elección")+
  theme_bw()
  
}

decision(Vj=6, Vi=5.5)
```

The probability of selecting alternative $j$ is $0.88$.

